ScPy: A SuperCollider Extension for Performing Numerical Computation via Embedded Python

### 2016-08-28 doc/paper.tex
some lit review + fixes

    SuperCollider~\cite{wilson2011supercollider}, a language for sound synthesis and algorithmic
    composition of audio, supports a wide range of synthesis, effect and analysis algorithms.
    However, available operations are limited to those implemented explicitly as Unit Generators
    (UGens). Since UGens are written in C/C++ and loaded as plugins during the SuperCollider server
    boot process, it is impossible to live code UGens, which limits the user to creating sound as a
    composition of existing UGens during a performance.  Many of the vector operations required for
    efficiently creating complex audio effects are notably missing or tedious to use. To overcome
    this, we present ScPy, a UGen which embeds Python within SuperCollider to enable the use of the
    optimized vector operations provided by the NumPy library.

@@

@@

ScPy is unlike many New Instruments for Musical Expression (NIMEs) because it does not have the
properties of any traditional instrument. However, it can still be considered a NIME because it
ultimately is a tool to enhance musical expression, just with an unconventional method of
interaction.  Dimension space analysis, as explored by Birnbaum et al.~\cite{birnbaum2005towards},
can facilitate the comparison of use cases for NIMEs.

\begin{figure}[h]
\caption{NIME dimension space analysis for ScPy.}
\vspace{1em}
\includegraphics[scale=0.2]{nime-plot}
\label{fig:nime}
\end{figure}

@@

@@

\begin{listing}[H]
    \begin{minted}[linenos=true]{SuperCollider}
(PyOnce("
    # comments in Python have different syntax!
    print('Hello from Python!')
    print('x is', x)
", (x: 62)))
    \end{minted}
    \caption{Hello world with ScPy.}
    \label{lst:helloworld}
\end{listing}

\begin{listing}[H]
    \inputminted[linenos=true]{SuperCollider}{../examples/template.sc}
    \caption{SuperCollider boilerplate for no-op FFT effect with ScPy.}
    \label{lst:boilerplate}
\end{listing}

@@

\begin{listing}[H]
    \begin{minted}[linenos=true]{python}
pv = PhaseVocoder(hop)

@@

\begin{table}[H]
    \caption{Type correspondence between SuperCollider and Python}
    \begin{center}
        \begin{tabular}{ll}
            \toprule
            SuperCollider type & Python type \\
            \midrule
            \texttt{Float} & \texttt{float} \\
            (subclass of) \texttt{UGen} & \texttt{ndarray} of \texttt{float} \\
            \texttt{Buffer} & \texttt{ndarray} of \texttt{float} \\
            \texttt{FFT} & \texttt{ndarray} of \texttt{complex} \\
            \texttt{Array} & \texttt{list} \\
            \bottomrule
        \end{tabular}
    \end{center}
\end{table}

@@

>>>>

    SuperCollider, a language for sound synthesis and algorithmic composition of audio, supports a
    wide range of synthesis, effect and analysis algorithms.  However, available operations are
    limited to those implemented explicitly as Unit Generators (UGens). Since UGens are written in
    C/C++ and loaded as plugins during the SuperCollider server boot process, it is impossible to
    live code UGens, which limits the user to creating sound as a composition of existing UGens
    during a performance.  Many of the vector operations required for efficiently creating complex
    audio effects are notably missing or tedious to use. To overcome this, we present ScPy, a UGen
    which embeds Python within SuperCollider to enable the use of the optimized vector operations
    provided by the NumPy library.

@@

\section{Literature Review}\label{sec:lit}

With any New Interface for Musical Expression (NIME), it is important to evaluate how it performs
and is interacted with across various dimensions. Birnbaum et al.~\cite{birnbaum2005towards}
present a new model for evaluating musical instruments by determining the position of the
instrument along 7 axes: \textit{Required Expertise}, \textit{Musical Control}, \textit{Feedback
Modalities}, \textit{Degrees of Freedom}, \textit{Inter-actors}, \textit{Distribution in Space},
and the \textit{Role of Sound}. Some of these axes represent a continuous spectrum, where as others
have discrete steps. These dimensions are arranged to form a heptagon in the 2D plane, where each
corner of the shape is placed at a distance to the center proportional to the score of the
instrument being evaluated along the corresponding axis. The specific layout of the dimensions used
in the paper additionally has properties which allow instruments of different classes to be easily
distinguished. Plots of instruments tend to occupy the right side of the graph, where as
installations tend to occupy the left. Although this model performs well on the instruments
evaluated in the paper, it is based on subjective evaluation and it is likely that it does not
adequately cover the space of all possible instruments.

ScPy is unlike many NIMEs because it does not have the properties of any traditional instrument.
However, it can still be considered a NIME because it ultimately is a tool for musical expression,
just with an unconventional method of interaction. A plot of ScPy in the 7-dimensional space
developed by Birnbaum et al.\ can be seen in~\autoref{fig:nime}.

\begin{figure}[ht]
\caption{NIME dimension space analysis for ScPy.}
\vspace{1em}
\includegraphics[scale=0.2]{nime-plot}
\label{fig:nime}
\end{figure}

The phase vocoder is an essential part of spectral audio manipulation. De G{\"o}tzen et
al.~\cite{de2000traditional} present an open source Matlab implementation along with a number of
tricks to make the implementation faster or simpler. The implementation is split into two
functions, \texttt{pv\_analyze} and \texttt{pv\_synthesize}. These functions operate as a tightly
coupled set due to the phase information maintained between frames. One of the techniques presented
is to perform a \texttt{fftshift} operation in order to simplify the phase relationship between
neighbouring bins. Another is to zero pad frames: by adding extra zeros on each side of a frame, a
larger FFT size can be used, producing better accuracy while maintaining good time resolution.

The code in this paper was used as a reference for the phase vocoder implementation in ScPy.
Additionally, some of the extra techniques mentioned could improve the quality or flexibility of
the implementation in the future.

Collins et al.~\cite{collins2003live} present an introduction to the field of live coding, where
code to generate sound is written live during a performance, often on a laptop computer. Although
the use graphical programming environments such as Reaktor, Pure Data, or MAX/MSP in live
performances is in some sense live coding, the paper focuses on the use of text based programming
environments. Many of the pros and cons of live coding are presented. Although live coding adds
some very interesting new possibilities to live music performance, it is not without it's
challenges. Case studies of two live coding environments are included: Slub and SuperCollider with
the just in time library.

ScPy can be used for live coding of audio effects, similarly to how the environments studied in
this paper can be used to live code synths and pattens. However, at the time of writing one issue
is blocking ScPy from realistically being used for live coding performances: the Python global
environment can be put into a state which prevents the code from executing without throwing an
exception, which could halt a performance in its tracks if the programmer made one of a number of
errors. This may be avoidable if more advanced error detection is implemented in the future.

Live coding presents an interesting contrast to contemporary music with many clear differences.
Nilson~\cite{nilson2007live} presents a review of the challenges of live coding
practice. While some argue that live coding must be performed live in front of an
audience, an alternate perspective suggests that time constraints are not necessary and that the
practice can refer to any improvisational music created by programming. The practice of live coding
is still quite young --- it is mentioned that professional violinists will typically have practiced
for 10000 hours, where as an email survey of live coders revealed most had dedicated about 100
hours. Of course, the prerequisite knowledge for live coding may be higher; many live coders have
extensive programming experience. Another thing mentioned is the high level of abstraction offered
by live coding creates a large startup time for producing meaningful output from a blank slate.
Additionally, many exercises are presented for improving one's live coding ability.

The ideas in Nilson's paper shed light on some of the challenges faced in this project.
One such challenge is the amount of writing required to produce an interesting result from a blank
slate. We intend to minimise this by providing many effect primitives and composition
mechanisms. However, it will be important to ensure the language is accessible to ensure it can be
quickly learned to a level of expertise where lack of technical knowledge is not an impediment.

While the Fourier transform is by far the most commonly used technique for spectral manipulation in
digital audio effects, there exist numerous other transformations which may be useful as well to
achieve certain audio effects. One such technique is the Mellin transform, described in a audio
effects context by De Sena and Rocchesso~\cite{de2004fast}. The Mellin transform has a scale
invariance property that is analogous to the shift invariance property of the Fourier transform. It
can be computed quickly by applying two extra steps before doing a Fourier transform: exponential
time warping and exponential multiplication. A few simple audio effects are presented using the
Mellin transform, time stretching, filtering and random phase deviation (whisperization).

As the Mellin transform can be implemented simply as an extension to the Fourier transform, it
should be fairly easy to add support for it in ScPy, and it will hopefully enable the use of
some different audio effects than can be achieved easily with the Fourier transform. However, the
paper mentions that a short-time Mellin transform has not been attempted, so some further work may
need to be done in that area first.

Domain-Specific Languages (DSLs) are small programming languages designed for use in a specific
problem domain. They often lack general purpose programming constructs and have a syntax which is
very concise for describing programs in their domain. Van Deursen et al.~\cite{van2000domain}
present a survey of available DSL literature, along with the pros and cons of using a DSL, example
DSLs, design methodology and implementation approaches. The biggest advantage of DSLs is that they
enable expression of solutions in a way that is common for the problem domain. However, they can be
costly to develop and difficult to design. Implementation techniques include the classical approach
of interpretation or compilation, embedded languages / domain-specific libraries, preprocessing, or
compiler extension. Additionally, the references of the paper each include a short summary.

This review reveals many of the possible options for the DSL used within the project. Although we
decided not to design a DSL for our project, analyzing the design and implementation techniques
presented here will be helpful for future work in this area.

Klingbeil~\cite{klingbeil2005software} describes the techniques used in a high quality graphical
spectral editing and resynthesis program called SPEAR\@. The FFT analysis is done taking into
account the fundamental frequency of the sound to analyze, so that the transform can be optimized
to accurately capture as many partials as possible. Spectral peaks are then selected and very quiet
peaks are discarded. A linear prediction method is used to follow the peaks over time. The data
structure for storage of spectral data involves division of time into a number of frames which keep
track of which partials are active during that time, so that the search for a specific partial can
be narrowed. User interface design choices are also covered in detail.

Although the interface and granularity of SPEAR differs significantly from those in project, the
techniques for high quality resynthesis presented should prove useful for improving the quality of
the sound output.

Keiler and Sylvain~\cite{keiler2002survey} present a survey of 6 techniques for accurate analysis
of stationary sounds. The first method, plain FFT, uses a Hann window and assumes the frequencies
of the sinusoids at centered within each bin. Parabolic interpolation notices that the shape of a
magnitude peak looks like a parabola, and thus attempts to fit a parabola to the magnitudes of bins
nearby each local maxima, yielding more accurate results than a plain FFT\@. The triangle algorithm
is similar to the parabolic interpolation method but fits a triangular function instead of a
parabola. Spectral reassignment uses the derivative of the analysis window to estimate the true
values of the bins as the center of gravity of the energy of each bin. The derivative algorithm
also uses the derivative of the analysis window to determine both a more accurate frequency and
amplitude. It can be further extended to use the first nth derivatives. The phase vocoder uses the
phase information of two successive FFTs to improve the frequency resolution by calculating the
phase derivative. Comparison of the techniques found that the spectral reassignment, derivative
algorithm, and phase vocoder methods had the best performance. Surprisingly, those three techniques
produced almost identical results, suggesting that they may have equivalent theoretical basis.

The two other techniques that performed similarly to the phase vocoder will be interesting to
analyze in more detail since they may present a solution to the \textit{phasiness} problem of the
phase vocoder due their lack of persistent state. However, they require extra processing so it is
not clear if the trade-off will be worthwhile.

Audio morphing is defined as a gradual change from one sound to the other. While morphing can be
achieved to some extent simply by cross fading two signals, perceptually this sounds simply like
one sound is becoming quieter as the other is becoming louder. Sethares and Bucklew
\cite{sethares2012kernel} propose density and kernel based techniques that aim to provide a more
convincing morph operation where there is no clear separation between the two sounds at any point
during the morphing process. The method allows the use of different kernel functions to provide
different kinds of morphing. These include \textit{Cross-fade}, \textit{Heat equation}, and
\textit{Harmonic integrity}. The morphing technique is implemented using a phase vocoder.

As a future extension to ScPy, these morphing techniques could be implemented in Python.

As Python is an easily extendible general purpose language, it has applications in many scientific
fields.  Glover et al.~\cite{glover2011python} present a discussion of the use of Python for audio
signal processing. They note that while audio processing libraries exist for many languages,
scripting languages such as Python or MATLAB see widespread adoption because of the speed at which
prototypes can be built. General purpose Python libraries such as NumPy, SciPy and Matplotlib
include many essential functions for audio processing, such as file IO, Fourier transforms, signal
processing and interpolation. There are additionally more specific libraries such as Simpl, a
sinusoidal modelling library, and Modal, an onset detection library.

Any of these libraries can easily be used within ScPy simply by installing and importing them.

Another open-source Python library which could be used in ScPy is librosa~\cite{mcfee2015librosa}
by McFee et al.\ librosa is a music signal analysis library designed to help transition music
information retrieval (MIR) researchers into using Python, as well as make MIR techniques readily
available to all Python programmers.\ librosa's core design choices and conventions are documented
in detail to inspire future library authors to follow similarly good coding practice. As well as
standard audio analysis functions, librosa also includes support for effects and audio IO.\@

librosa has quite an extensive list of features and would likely cover all of the audio related
functionality which is missing from NumPy. It can be used with ScPy simply by installing and importing
it.

The ixi lang~\cite{magnusson2011ixi} is a live coding language. It is implemented as an interpreter
built in SuperCollider, a choice which allows for SuperCollider and ixi lang code to be mixed
together in a similar manner to ScPy. The ixi lang is built to reduce the startup time in a live
coding performance and enable faster musical experimentation through syntax more concise than
SuperCollider. The paper also contains user feedback on the usability of the ixi lang.

Since the ixi lang is similar to ScPy in many ways, many of the design choices made in ixi lang are
also appropriate in ScPy. Additionally, lessons learned from the user feedback for ixi lang could
potentially be used to improve usability of ScPy.

@@

@@

In order to facilitate the use of ScPy in live coding performances, many errors are ignored or
handled in a way that can be easily recovered from. In particular, floating point errors such as
division by zero or other invalid operations are ignored. Special values such as infinity or NAN
are replaced with zeros. This is done because SuperCollider has poor handling of these special
values and in some cases will crash if they are found. Python exceptions are printed to the
SuperCollider post window, where they can easily be examined and corrected by the user.

@@

\begin{listing}[ht]
    \begin{minted}[linenos=true]{SuperCollider}
(PyOnce("
    # comments in Python have different syntax!
    print('Hello from Python!')
    print('x is', x)
", (x: 62)))
    \end{minted}
    \caption{Hello world with ScPy.}
    \label{lst:helloworld}
\end{listing}

\begin{listing}[ht]
    \inputminted[linenos=true]{SuperCollider}{../examples/template.sc}
    \caption{SuperCollider boilerplate for no-op FFT effect with ScPy.}
    \label{lst:boilerplate}
\end{listing}

@@

\begin{listing}[ht]
    \begin{minted}[linenos=true]{python}
pv = PhaseVocoder(hop)

@@

\begin{table}[ht]
    \caption{Type correspondence between SuperCollider and Python}
    \begin{center}
        \begin{tabular}{ll}
            \toprule
            SuperCollider type & Python type \\
            \midrule
            \texttt{Float} & \texttt{float} \\
            (subclass of) \texttt{UGen} & \texttt{ndarray} of \texttt{float} \\
            \texttt{Buffer} & \texttt{ndarray} of \texttt{float} \\
            \texttt{FFT} & \texttt{ndarray} of \texttt{complex} \\
            \texttt{Array} & \texttt{list} \\
            \bottomrule
        \end{tabular}
    \end{center}
\end{table}

@@

TODO data flow diagram
